---
title: Diversity calculation
author: "Masatoshi Katabuchi"
date: "November 20, 2025"
fontsize: 12pt
format:
  html:
    theme: coderpro
    highlight-style: github
    toc: true
    toc-depth: 2
    number-sections: true
    smooth-scroll: true
    embed-resources: false
    include-after: |
      <link rel="stylesheet" href="assets/fontawesome-free-7.1.0-web/css/all.min.css">
      <script src="assets/fontawesome-free-7.1.0-web/js/all.min.js"></script>
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      warning=FALSE,
                      cache=TRUE,
                      message=FALSE)
```

<a href="https://bsky.app/profile/mattocci.bsky.social"><i class="fa-brands fa-bluesky fa-1x"></i> Bluesky</a>
<a href="https://github.com/mattocci27/"><i class="fa-brands fa-github fa-1x"></i> Github</a>
<a href="mailto:mattocci27@gmail.com"><i class="fa-solid fa-envelope fa-1x"></i> Email</a>

Course materials for AFEC 2025 at XTBG.

# Objectives

- Prepare species $\times$ site matrices and trait data from csv files.
- Calculate diversity indices.

# Prerequisites

You can work through these exercises with a standard R installation; just grab the files and install a few packages.

1. Clone the repository (or download and unzip it):

   ```{bash, eval=FALSE}
   git clone https://github.com/mattocci27/phy-fun-div.git
   ```

2. (**Optional**) Restore the environment with `renv`. This ensures identical package versions but takes longer the first time.

   ```{r, eval=FALSE}
   renv::restore()
   ```

3. Prefer a quicker start? Install the required packages manually:

   ```{r, eval=FALSE}
   install.packages(c(
     "tidyverse", "picante", "FD",
     "DT", "here", "vegan", "ape"
   ))
   ```

If you want to skip `renv`, comment out the autoload lines in `.Rprofile`.

```{r}
# Sys.setenv(RENV_DOWNLOAD_METHOD = "curl")
# Sys.setenv(RENV_CONFIG_PAK_ENABLED = "TRUE")
# options(renv.install.staged = FALSE)

# if (file.exists("renv/activate.R")) {
#   source("renv/activate.R")
#   if ("renv" %in% loadedNamespaces()) {
#     renv::settings$use.cache(TRUE)
#   }
# }
```

# Load packages

```{r}
library(tidyverse)
library(picante)
library(FD)
```

# Data

## Community

First, we import the community data.

```{r}
d <- read_csv("data/samp.csv")
```

### Using a relative path

- I prefer `read_csv` but `read.csv` is fine too.
- Our working directory is `phy-fun-div`, and the relative path to `samp.csv` is `data/samp.csv`.

The directory structure looks like this:

```
.
├── data/
│   ├── dummy_trait.csv
│   ├── dummy_tree.newick
│   ├── samp.csv
│   └── soil.csv
```

### Using `here` package

We can also use `here` package to specify the absolute path to the file.
This is useful when you are working in a sub-directory.
For example, you might be in `docs/` directory, because your qmd file is there, but `here` will help you to find the correct absolute path to the data file.

```{r, error=TRUE}
# (For demonstration only — normally you shouldn't change working directories)
setwd("docs")
getwd()
read_csv("data/samp.csv")
```

```{r}
here::here("data/samp.csv")
```

```{r}
d <- read_csv(here::here("data/samp.csv"))
setwd("..")
```

### Avoiding absolute paths

Avoid using absolute paths like this:

```{r, eval=FALSE}
d <- read_csv("/Users/mattocci/Dropbox/5-Tools/ghq/github.com/mattocci27/phy-fun-div/data/samp.csv")
```

Absolute paths are different on different computers (lack of reproducibility).
It is also uncomfortable to make file system structure public (security risk).

### Community data

Now let's look at the community data.

```{r}
d
DT::datatable(d)
```

Then, we want to make a species $\times$ site matrix.
`tapply` is a useful function here.

```{r}
tapply(d$abund, d$Species, sum)
```

```{r}
samp <- tapply(d$abund, list(d$Site, d$Species), sum)
samp
class(samp)
```

## Phylogeny

```{r, fig.height=20}
phylo <- read.tree("data/dummy_tree.newick")
plot(phylo)
```

## Traits

| Abbreviation | Trait                                      | Unit              |
| ------------ | ------------------------------------------ | ----------------- |
| LMA          | Leaf mass per area                         | g m^-2^           |
| LL           | Leaf lifespans (longevity)                 | months            |
| Amass        | Maximum photosynthetic rates per unit mass | nnoml g^-1^ s^-^1 |
| Rmass        | Dark respiration rates per unit mass       | nnoml g^-1^ s^-^1 |
| Nmass        | Leaf nitrogen per unit mass                | %                 |
| Pmass        | Leaf phosphorus per unit mass              | %                 |
| WD           | Wood density                               | g cm^-3^          |
| SM           | Seed dry mass                              | mg                |


```{r}
trait <- read_csv("data/dummy_trait.csv")

DT::datatable(trait)
```

## Check how traits look like first

```{r}
trait_long <- trait |>
  pivot_longer(lma:sm, names_to = "trait")

trait_long

ggplot(trait_long, aes(x = value)) +
  geom_histogram(position = "identity") +
  facet_wrap(~ trait, scale = "free") +
  theme_bw()
```


Probably we can do log-transformation for all the traits except for WD.

```{r}
trait2 <- trait |>
  mutate(
    across(c(lma, ll, a_mass, r_mass, n_mass, p_mass, sm),
       ~ log(.),
        .names = "log_{.col}")) |>
  dplyr::select(sp, log_lma, log_ll, log_a_mass, log_r_mass, log_n_mass, log_p_mass, wd, log_sm)

trait2 |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  DT::datatable()
```

```{r}
trait2 |>
  pivot_longer(log_lma:log_sm, names_to = "trait") |>
  ggplot(aes(x = value)) +
  geom_histogram(position = "identity") +
  facet_wrap(~ trait, scale = "free") +
  theme_bw()
```

# First-order metrics (without phylogeny or traits)

## Species richness

```{r}
samp > 0
apply(samp > 0, 1, sum)
```

## Shannon

$H' = - \sum_i^n p_i\mathrm{log}p_i$, where $p_i$ is the relative abundance for
species *i*.

```{r}
shannon <- function(abund) {
  p0 <- abund / sum(abund)
  p <- p0[p0 > 0]
  -sum(p * log(p))
}
```

```{r}
apply(samp, 1, shannon)
```

You don't have to reinvent the wheel.

```{r}
vegan::diversity(samp, index = "shannon")
```

## Nonmetric Multidimensional Scaling (NMDS)

```{r}
res_mds <- metaMDS(samp)

plot(res_mds)
```

We can use the function `ordiplot` and `orditorp` to add text to the plot in place of points to make some more sense.

```{r}
ordiplot(res_mds, type = "n")
orditorp(res_mds, display = "species", col = "red", air = 0.01)
orditorp(res_mds, display = "sites", cex = 1.25, air = 0.01)
```

# Phylogenetic metrics

## Branch length based metric

### PD

```{r}
res_pd <- pd(samp, phylo)
res_pd
```


You can always see the help.

```{r, eval=FALSE}
?pd
```


## Distance based metric

`cophenetic()` creates distance matrices based on phylogenetic trees. Let's see the first 5 species.

```{r}
cophenetic(phylo)[1:5, 1:5]
```


### MPD

$MPD = \frac{1}{n} \Sigma^n_i \Sigma^n_j \delta_{i,j} \; i \neq j$, where
$\delta_{i, j}$ is the pairwise distance between species *i* and *j*

```{r}
res_mpd <- mpd(samp, cophenetic(phylo))
res_mpd
```

The above vector shows MPD for each site.

### MNTD

$MNTD = \frac{1}{n} \Sigma^n_i min \delta_{i,j} \; i \neq j$, where $min \delta_{i, j}$ is the minimum distance between species *i* and all other species in the community.

```{r}
res_mntd <- mntd(samp, cophenetic(phylo))
res_mntd
```



# Functional metrics

## Community weighted means (CWM)

$$
\mathrm{CWM}_i = \frac{\sum_{j=1}^n a_{ij} \times t_{j}}{\sum_{j=1}^n a_{ij}}
$$

or

$$
\mathrm{CWM}_i = \frac{\boldsymbol{a_i} \cdot \boldsymbol{t}}{\sum_{j=1}^n a_{ij}},
$$

where $a_{ij}$ is the abundance of species *j* in community *i*, and $t_{j}$ is a trait value of species *j*.

```{r}
tmp <- trait2 |>
  filter(sp %in% colnames(samp))

tmp

(ab <- apply(samp, 1, sum))
# %*% denotes inner product
(cws <- samp %*% as.matrix(tmp[, -1]))
```
```{r}
(cwm <- cws / ab)
```

The species $\times$ site matrix and the species $\times$ trait matrix became the trait $\times$ site matrix.

## Distance based metrics

### Prepare a trait distance matrix

We have a `data.fame` or `tibble` object of traits.
First, we need to prepare a trait matrix, then a distance matrix based on trait values.

```{r}
trait_mat0 <- as.matrix(trait2[, -1])
rownames(trait_mat0) <- trait2$sp
```

Let's see a subset of the trait matrix

```{r}
trait_mat0[1:5, 1:5]
```

Then, we will make a trait distance matrix based on the Euclidean distance.
There are other distance measures, for example [Gower's Distance](https://www.rdocumentation.org/packages/gower/versions/0.2.1/topics/gower_dist), but we focus on the Euclidean distance today.

Before calculating distance, we need to make sure the unit change in distances is the same for different traits.
We will scale trait values so that they have mean = 0 and SD = 1.
(e.g., $(X_i - \mu) / \sigma$)

```{r, fig.height=6}
# (trait_mat0 - mean(trait_mat0)) / sd(trait_mat0)
trait_mat <- scale(trait_mat0)

par(mfrow = c(2, 2))
hist(trait_mat0[, "log_lma"])
hist(trait_mat[, "log_lma"])
hist(trait_mat0[, "wd"])
hist(trait_mat[, "wd"])
par(mfrow = c(1, 1))
```

Now we can make a trait distance matrix.

```{r}
trait_dm <- as.matrix(dist(trait_mat))
```

Let's see the first 5 species.

```{r}
trait_dm[1:5, 1:5]
```

### MPD (Mean Pairwise Distance)

```{r}
mpd(samp, trait_dm)
ses.mpd(samp, trait_dm)
```

**Effect size (ES)**

$$
ES = \frac{\bar{x_1} - \bar{x_2}}{\sigma} \sim Normal(\bar{x_1} - \bar{x_2}, 1)
$$

where $\bar{x_1}$ is the mean value of $x_1$, $\bar{x_2}$ is the mean value of $x_2$, and $\sigma$
is the standard deviation of the pooled data.

```{r, echo=FALSE}
xx2 <- seq(-3, 3, length = 80)
yy2 <- dnorm(xx2, 0, 1)
plot(xx2, yy2, type = "l",
     main = "Normal(0, 1)",
     xlab = "X", ylab = "Density")
```

In null model analyses, we can translate this into a standardized effect size
(SES):

$$
SES = \frac{obs - \bar{rand}}{\sigma_{rand}}
$$

where obs is the observed metric, $\bar{rand}$ is the mean value of the metric in
null communities, and $\sigma_{rand}$ is the standard deviation of the metric in
the null communities.

### MNTD (Mean Nearest Taxon Distance)

```{r}
mntd(samp, trait_dm)
ses.mntd(samp, trait_dm)
```

## Branch length based metric

### FD

We will make a functional dendrogram using clustering methods.
We use [UPGMA](https://en.wikipedia.org/wiki/UPGMA) in this example.

```{r, fig.width=15, fig.height=15}
t_clust <- hclust(dist(trait_mat), method = "average")

plot(t_clust)
```


### More functional diversity metrics

```{r, eval=TRUE}
res_fd <- dbFD(trait_mat[colnames(samp), ], samp)
res_fd
```

<iframe src="appendixA.htm" width="900" height="1200" allowfullscreen frameborder="0"></iframe>

# Related

- Aiba, M., Katabuchi, M., Takafumi, H., Matsuzaki, S.S., Sasaki, T. and Hiura, T. (2013), Robustness of trait distribution metrics for community assembly studies under the uncertainties of assembly processes. Ecology, 94: 2873-2885. [doi:10.1890/13-0269.1](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0269.1)

- [Phylogenetic diversity in R](https://daijiang.name/en/2014/05/04/notes-func-phylo-book-1/)

- [Functional diversity in R](https://daijiang.name/en/2014/05/11/functional-diversity-in-r/)

- [Biodiversity analysis in R](https://www.kembellab.ca/r-workshop/biodivR/SK_Biodiversity_R.html)

- [Measuring functional diversity in R](http://traits-dgs.nceas.ucsb.edu/workspace/r/r-tutorial-for-measuring-functional-diversity/Measuring%20Functional%20Diversity%20in%20R.pdf/attachment_download/file)


```{r}
# sessionInfo()
```
